Thesis - Upwelling and MHW/Cold spells analyses

### 2019-09-17 Write_up/write_up.Rmd
update

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundances, diversity, and distributions of marine organisms, and moreover, critically affect marine ecosystem processes (Add in references). Globally, upwelling occurs throughout the world’s oceans, however four major eastern boundary upwelling systems (EBUS) are recognised, each considerably impacting coastal systems. These respective systems are characterized as being vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun ; @pauly]. Eastern boundary upwelling regions only cover 1% of the world’s ocean surface but are among the most productive regions of the world.

@@

The dynamics of upwelling within the nearshore (<400m from the coastline) remains poorly understood.. Upwelling metrics are often calculated far offshore because measurements are typically taken from large ships that cannot move close to the shore (Reference). Cold temperatures are important in describing species population distribution limits and are proven to set limits in the north or southwards higher latitudes (Firth et al., 2011). There is by no means an understanding of how the upwelling process varies at local versus broader scales. We hypothesize that upwelling is more dominant within coastal regions instead of broader scale. Very few studies have addressed the changes in the intensity or duration of cold temperature spells or cold events at varying distances from the coast within the Benguela region.

The central assumption of this project was that remotely-sensed SST data and *in situ* seawater temperature data may be compared. Remotely sensed SST data and *in situ* seawater temperature data were obtained at different resolutions and so slight changes may be found within the different datasets. The primary aim of the first study was to make use of various remotely sensed SST and in-situ collected coastal seawater temperature data. The most appropriate dataset for the measurement of the nearshore temperatures within South Africa was the South African Coastal Temperature Network (SACTN) dataset (Smit et al. 2013). It was useful to examine whether or not the same upwelling patterns exist at different distances from the coastline and if these patterns that exist are evident within the various datasets. This investigation examines recent trends and variation of upwelling events at different distances from the coastline.

@@

In order to compare abiotic variables such as wind action and direction along the South African coastline, large historical datasets for temperature and wind were analysed and accessed.

@@

This research study lies within the Benguela region along the west coast of South Africa. Our study sites covered the full range of this coastal region (Fig.1). The South African Coastal Temperature Network (SACTN) data serves as the *in situ* temperature dataset used in this study. This dataset consists of coastal seawater temperatures obtained from 129 sites along the South African coastline which is approximately 3000km long and is distinct in that it is bordered by the Benguela and Agulhas Currents (e.g.,Roberts 2005; Hutchings et al. 2009). These currents are known to largely influence the seawater temperature and marine life along the South African coastline. The SACTN has collected SST data along this unique coastline from as early as 1972 till present date, with contributions made from several organisations and govermental departments. As a result, temperature within this datasets were obtained by using different methods and instruments, such as hand held thermometers or Underwater temperature reorders( UTRs). The precision of the instruments used vary accordingly, with hand-held thermometers recording temperature at a precision of 0.5°C as opposed to the underwater temperature recorders which have an instrument precision of 0.001°C. The duration and extent of recordings per sites were uneven with the longest time series in the dataset being that of Gordons Bay which started on 13 September 1972 and concluded on 26 January 2017.  During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 entries were added during the 1990s, and 18 entries were added during the 2000s. Recordings are still ongoing at many of these sites.

@@

An obvious previledge exist when making using of *in situ* data over satellite data as they provide a more accurate representation of the thermal properties closer to the coast where satellite temperatures do not capture temperature properties well. Smit et al (2013) have shown that satellite data collected along the South African coastline represent a warm biases reaching as high as 6°C over *in situ* temperatures within the nearshore environment.

@@

Wind action was an important variable in this study as it was hypothesised that wind direction and wind speed may have a direct influence on the intensity and duration of upwelling events; consequently, wind action was investigated for its impact on upwelling at specific sites along the west coast of South Africa. Wind data was obtained from the South African Weather Service (SAWS), and were provided at three-hour resolutions. Specific wind characteristics were measured, namely, wind direction (dirw) and wind speed (spw).

At a broad scale, the South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions, each with contrasting temperatures [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings] which provides a natural labratory for this study. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflect a distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. This region represents monthly mean temperatures at generally intermediate at cold and warm temperate [@luning]. In order to examine upwelling patterns between sites, the appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites with a time series of longer than 30 years. This yield a total of five sites within the Benguela region.

@@

Inorder to determine whether or not upwelling signals are present at different distances from the coastline and to test which dataset shows clear, visible representation of these events, it was necessary to define when upwelling is occuring. For this research study we aimed to determine a threshold value for which we could identify that upwelling was occuring at a particular time. Given that upwelling is primarily caused by alongshore, equator ward winds, it was necessary to consider temperature and wind values when determining a threshold for temperature to be experiencing an upwelling event. We then decided to make use of the following equation $$ UpwellingIndex = μ{(Cosθ − coastal angle)} $$

In this equation μ represents the wind speed (m/s) and θ represents the wind direction which is measured in degrees. The coastal angle is the angle of each of the sites perpendicular to the coastline. Thereafter, the 'exceedance' function within the heatwaveR package was used to determine the consecutive number of days at or above which the upwelling index value is ment to be. Here it was specified that the minimum duration for upwelling is 1 day with a threshold value obtained from the upwelling indeces as 1. Inorder to obtain a more approved threshold value, a percentile value was determined. To determine changes within the upwelling values, the 'detect event' fuction within the heatwaveR package was used. Thereafter, to determine the upwelling event metrics ( frequency and intensity) the detect event function was used as a first filter with a threshold of 25th percentile. Values obtained from the upwelling indeces were used as a second threshold. An R script (https://github.com/AmierohAbrahams/Upwelling_MCS) was used to further used to calculate the individual upwelling events for both the *in situ* and remotely sensed SST data. Because upwelling events were calculated relative to percentile exceedences, rather than an absolute definition such as periods with temperatures above a fixed temperature threshold, upwelling events occured any time of the year with upwelling shown to be more dominant during spring and summer months.

>>>>

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundances, diversity, and distributions of marine organisms, and moreover, critically affect marine ecosystem processes (Add in references). Globally, upwelling occurs throughout the world’s oceans, however four major eastern boundary upwelling systems (EBUS) are recognised, each considerably impacting coastal systems. These respective systems are characterized as being vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun ; @pauly]. The oceanography of the inshore waters off the west coast of South Africa is dominated by an upwelling system which is regarded along with the Peru, California and Humboldt current as Eastern boundry upwelling regions [@andrews]. Eastern boundary upwelling regions only cover 1% of the world’s ocean surface but are among the most productive regions of the world. As such there has been a constant interest in upwelling systems. Major upwelling programs were carried through off off Peru (e.g. WALSH. KELLEY, DUGDALE and FROST, 1971) and North-West Africa (e.g. BARBER, 1977).

@@

The dynamics of upwelling within the nearshore (<400m from the coastline) remains poorly understood. Upwelling metrics are often calculated far offshore because measurements are typically taken from large ships that cannot move close to the shore [@andrews]. Cold temperatures are important in describing species population distribution limits and are proven to set limits in the north or southwards higher latitudes (Firth et al., 2011). There is by no means an understanding of how the upwelling process varies at local versus broader scales. We hypothesize that upwelling is more dominant within coastal regions instead of broader scale. Very few studies have addressed the changes in the intensity or duration of cold temperature spells or cold events at varying distances from the coast within the Benguela region.

The central assumption of this project was that remotely-sensed SST data and *in situ* seawater temperature data may be compared. Remotely sensed SST data and *in situ* seawater temperature data were obtained at different resolutions and so slight changes may be found within the different datasets. The primary aim of the first study was to make use of various remotely sensed SST and in-situ collected coastal seawater temperature data. The most appropriate dataset for the measurement of the nearshore temperatures within South Africa was the South African Coastal Temperature Network (SACTN) dataset [@smit]. It was useful to examine whether or not the same upwelling patterns exist at different distances from the coastline and if these patterns that exist are evident within the various datasets. This investigation examines recent trends and variation of upwelling events at different distances from the coastline.

@@

At a broad scale, the South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions, each with contrasting temperatures [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings] which provides a natural labratory for this study. Annual mean coastal seawater temperature within this region represents a coastal seawater temperature range of 12.3 ± 1.2°C at the weastern limit near the Namibian border. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflect a distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. This region represents monthly mean temperatures at generally intermediate at cold and warm temperate [@luning].  In order to accurately examine upwelling patterns at various distances from the coast several sites were selected as this provides a more accurate representation, the appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites with a time series of longer than 30 years. This yield a total of five sites within the Benguela region (Fig 1).

@@

The South African Coastal Temperature Network (SACTN) data serves as the *in situ* temperature dataset used in this study. This dataset consists of coastal seawater temperatures obtained from 129 sites along the South African coastline. The SACTN has collected SST data along the unique coastline of South Africa from as early as 1972 till present date, with contributions made from several organisations and govermental departments. As a result, temperature within this datasets were obtained by using different methods and instruments, such as hand held thermometers or Underwater temperature reorders( UTRs). The precision of the instruments used vary accordingly, with hand-held thermometers recording temperature at a precision of 0.5°C as opposed to the underwater temperature recorders which have an instrument precision of 0.001°C. The duration and extent of recordings per sites were uneven with the longest time series in the dataset being that of Gordons Bay which started on 13 September 1972 and concluded on 26 January 2017.  During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 entries were added during the 1990s, and 18 entries were added during the 2000s. Recordings are still ongoing at many of these sites.

@@

An obvious previledge exist when making using of *in situ* data over satellite data as they provide a more accurate representation of the thermal properties closer to the coast where satellite temperatures do not capture temperature properties well and thus can more accurately explain upwelling within the coastal inshore environment. Smit et al (2013) have shown that satellite data collected along the South African coastline represent a warm biases reaching as high as 6°C over *in situ* temperatures within the nearshore environment.

@@

Wind action was an important variable in this study as it was hypothesised that wind direction and wind speed may have a direct influence on upwelling; consequently, wind action was investigated for its impact on upwelling at specific sites along the west coast of South Africa. Wind data was obtained from the South African Weather Service (SAWS), and were provided at three-hour resolutions. Specific wind characteristics were measured, namely, wind direction (dirw) and wind speed (spw). Since the wind data were modelled at three-hour resolutions, they were converted into daily data points in order to compare them with the temperature data. The circular() function in R software was then used to create circular objects around the wave data in order to calculate the daily wind parameters. ERA5 hourly data on pressure levels was used as a second wind source. This dataset comprise of u and v variables from 1973 to present.

With temperature and wind values now corresponding to their respective sites and dates, the hypothesis testing upwelling patterns at different distances from the coastline could now be tested. To do this, it was important to identify upwelling.

@@

Inorder to determine the change in upwelling signals at different distances from the coastline and to test which dataset shows clear, visible representation of these events, it was first necessary to define when upwelling was occuring. We then aimed to determine a threshold value for which we could identify when upwelling was occuring within this region. Given that upwelling is primarily caused by alongshore, equator ward winds, it was necessary to consider both temperature and wind values. To determine the upwelling condition, an upwelling index was calculated using the formula presented in Fielding and Davis (1989). The upwelling index provided an indication of the intensity of the upwelling process taking place using the following equation:

                  $$ UpwellingIndex = μ{(Cosθ − coastal angle)} $$

In this expression, μ represents the wind speed (m/s) and θ represents the wind direction which is measured in degrees. The coastal angle is the angle of each of the sites perpendicular to the coastline. The index relies heavily on wind speed and direction data inorder to determine the intensity of upwelling likely to occur. Thereafter, the 'exceedance' function within the heatwaveR package was used to determine the consecutive number of days at or above which the upwelling index value is ment to be. Here it was specified that the minimum duration for upwelling is 1 day with a threshold value obtained from the upwelling index as 1. Inorder to obtain a more approved threshold value, a percentile value was determined. To determine changes within the upwelling values, the 'detect event' fuction within the heatwaveR package was used. Thereafter, to determine the upwelling event metrics ( frequency and intensity) the detect event function was used as a first filter with a threshold of 25th percentile. Values obtained from the upwelling indeces were used as a second threshold. An R script (https://github.com/AmierohAbrahams/Upwelling_MCS) was used to calculate the individual upwelling events for both the *in situ* and remotely sensed SST data. Because upwelling events were calculated relative to percentile exceedences, rather than an absolute definition such as periods with temperatures above a fixed temperature threshold, upwelling events occured any time of the year with upwelling shown to be more dominant during spring and summer months. For the use of the heatwaveR package it was best to use a time series of 30years.
